Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2020-04-09 article.tex
Hypotheses

For MHC-1, it was found that epitopes derived from transmembrane helices (TMHs)
are over-presented by all human leukocyte antigen (HLA)-A and most HLA-B super 
types [\cite{bianchi2017transmembrane}]. One explanation is that the 
presentation of especially TMHs may have an evolutionary advantage for 
the (human) host. The argumentation for this is that the TMHs are more 
conserved due to the functional requirement of being able to span a lipid 
bilayer. Additionally, there are limited escape mutations possible for the 
pathogen, as many will result in a non-functional TMH.

@@

\begin{itemize}
  \item $\mathcal{H}_1$: MHC-I has the same percentage of epitopes overlapping
    (that is 5.73 to 5.59 percent, with a confidence interval of 99 percent) 
    with TMHs in COVID-19 as in 
4 + ((117 / 135) * 2) = 5.73
4 + ((107 / 135) * 2) = 5.59

  \item $\mathcal{H}_2$: MHC-II binds as often to TMHs as the cytosolic and extracellular
    parts of a pathogen's transmembrane proteins
  \item $\mathcal{H}_3$: The TMHs of [a pathogen] have the same mutation rate 
    as the cytosolic and extracellular
    parts of its transmembrane proteins
\end{itemize}

@@

Transmembrane helices and strong MHC-II-binding peptides
were predicted for a tuberculosis reference proteome 
from \url{https://www.ebi.ac.uk/reference_proteomes}
using the \verb;epitopeome; 
\richel{
  I'd enjoy a better name
}
R package [\cite{epitopeome}].
We picked the three MHC-II alleles that are most abundant 
in the current human population, 
which are DRB4*0101, HLA-DPA10103-DPB10402 
\richel{
  We decided for DPB1*0402, but there are multiple alleles matching.
  I just picked the first one of the alleles matching, which are:
  HLA-DPA10x-DPB10402, where x equals 103-110, 201-204, 301-303 or 401
}
and DQA1*0501/DQB1*0301 \richel{reference here}.
The 5\% peptides with the lowest IC50 values were defined as binders.
We then simply counted the number of amino acids that were present inside the
cell, within the membrane or outside of the cell, as well as if it was part 
of a strong MHC-II binding site, as shown in figue \ref{tab:results}.

The \verb;epitopeome; R package \cite{epitopeome} binds \richel{pun intended} 
together the \verb;tmhmm; [\cite{tmhmm}] and \verb;netmhc2pan; 
[\cite{netmhc2pan}] 
R packages. \verb;tmhmm; provides an R interface to
TMHMM [\cite{krogh2001predicting}, \cite{sonnhammer1998hidden}], a tool
to predict where membrane proteins' amino acids are located within the
membrance.   
\verb;netmhc2pan; provides an R interface to
NetMHC2pan [\cite{jensen2018improved}], a tool
to predict MHC-II binding to proteins.

@@

Using the \verb;tmhprot; R package [\cite{tmhprot}], the DNA alignment is split
into two alignments, one for the TMH parts, another for the non-TMH
parts. Each alignment was tested by the \verb;mcbette; 
R package [\cite{mcbette}] to select the Bayesian inference model with
the highest evidence (a.k.a. the marginal likelihood) using the nested
sampling approach as described in \cite{maturana2018model},
using the popular Bayesian phylogenetic tool 
BEAST2 [\cite{bouckaert2014beast}] in the back-end.
\verb;mcbette; used a set of 40 candidate inference models, 
consisting of all combinations of 
4 site models (JC, HKY, TN, GTR), 
2 clock models (strict, RLN) and 
5 tree priors (Yule, BD, CBS, CCP, CEP).

For each alignment, the inference model with the highest evidence
is used in a Bayesian inference. From each Bayesian posterior,
the parameter estimates regarding mutation rates (including clock rate)
were obtained and compared to determine which realized mutation rate is lower.
The Markov chain Monte Carlo was set up in such a way that the effective sample
size for the likelihood of the inference model is above the recommended value
of 200 [\cite{bouckaert2014beast}].

@@

>>>>

For MHC-1, it was found that epitopes derived from transmembrane helices (TMHs)
are over-presented by all human leukocyte antigen (HLA)-A and 
most HLA-B super types (\cite{bianchi2017}). 
One explanation is that the 
presentation of especially TMHs may have an evolutionary advantage for 
the (human) host. The argumentation for this is that the TMHs are more 
conserved due to the functional requirement of being able to span a lipid 
bilayer. Additionally, there are limited escape mutations possible for the 
pathogen, as many will result in a non-functional TMH.

@@

\begin{itemize}
  \item $\mathcal{H}_1$: MHC-I has the same percentage of epitopes overlapping
    with TMHs in Homo sapiens as in COVID-19
  \item $\mathcal{H}_2$: MHC-II has the same percentage of epitopes overlapping
    with TMHs as expected by chance
  \item $\mathcal{H}_3$: The TMHs of COVID-19 have the same mutation rate 
    as the cytosolic and extracellular parts of its transmembrane proteins
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{MHC-I}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We used the same analysis as \cite{bianchi2017},
except that instead of a human reference proteome,
we used the proteome of the first sequenced COVID-19 strain (\cite{wu2020new}),
with a GenBank ID of MN908947.3, as can be downloaded 
from \url{https://www.ncbi.nlm.nih.gov/nuccore/MN908947}.

Bianchi and colleagues obtained a distribution of 
percentages of MHC-I epitopes overlap with TMHs in Homo sapiens
for the different HLA haplotypes, with has an average of 5.3\%.
We obtained a similar distribution of percentages of MHC-I epitopes that 
overlap with TMHs for the different HLA haplotypes, but then applied to
COVID-19.

We compare these distribution using a two-sample 
Kolmogorov-Smirnov (KS) test. The KS test determines if two samples
are derived from a same distribution, without making assumptions
regarding the shape of that distribution. We will reject
the null hypothesis that MHC-I has the same percentage of epitopes 
overlapping with TMHs in Homo sapiens as in COVID-19 when 
the KS statistic follows the relationship as show in 
figure \ref{fig:ks}, for an $\alpha$ of $0.05$.
is below

\begin{figure}[!htbp]
  \includegraphics[width=0.4\textwidth]{pics/ks.png}
  \caption{
    Condition of a Kolmogorov-Smirnov (KS) statistic to rejects a null
    hypothesis. $D_{n,m}$ is the KS statistic, $n = m$ is the size
    of the samples, in this case, the number of HLA haplotypes ($n = m = 13$).
    $\alpha = 0.05$ is the significance level.
  }
  \label{fig:ks}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{MHC-2}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Transmembrane helices and strong MHC-II-binding peptides
were predicted for a tuberculosis reference proteome 
from \url{https://www.ebi.ac.uk/reference_proteomes}
using the \verb;epitopeome; R package (\cite{epitopeome}).
We picked the three MHC-II alleles that are most abundant 
in the current human population, 
which are DRB4*0101, HLA-DPA10103-DPB10402 
\richel{
  We decided for DPB1*0402, but there are multiple alleles matching.
  I just picked the first one of the alleles matching, which are:
  HLA-DPA10x-DPB10402, where x equals 103-110, 201-204, 301-303 or 401
}
and DQA1*0501/DQB1*0301 \richel{reference here}.
The 5\% peptides with the lowest IC50 values were defined as binders.
We then simply counted the number of amino acids that were present inside the
cell, within the membrane or outside of the cell, as well as if it was part 
of a strong MHC-II binding site, as shown in figue \ref{tab:results}.

The \verb;epitopeome; R package (\cite{epitopeome}) binds 
together the \verb;tmhmm; (\cite{tmhmm}) and \verb;netmhc2pan; 
(\cite{netmhc2pan}) 
R packages. \verb;tmhmm; provides an R interface to
TMHMM (\cite{krogh2001predicting, sonnhammer1998hidden}), a tool
to predict where membrane proteins' amino acids are located within the
membrance.   
\verb;netmhc2pan; provides an R interface to
NetMHC2pan (\cite{jensen2018improved}), a tool
to predict MHC-II binding to proteins.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evolutionary conservation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Using the \verb;tmhprot; R package (\cite{tmhprot}), the DNA alignment is split
into two alignments, one for the TMH parts, another for the non-TMH
parts. Each alignment was tested by the \verb;mcbette; 
R package (\cite{mcbette}) to select the Bayesian inference model with
the highest evidence (a.k.a. the marginal likelihood) using the nested
sampling approach as described in \cite{maturana2018model},
using the popular Bayesian phylogenetic tool 
BEAST2 (\cite{bouckaert2014beast}) in the back-end.
\verb;mcbette; used a set of 40 candidate inference models, 
consisting of all combinations of 
4 site models (JC, HKY, TN, GTR), 
2 clock models (strict, RLN) and 
5 tree priors (Yule, BD, CBS, CCP, CEP).

For each alignment, the inference model with the highest evidence
is used in a Bayesian inference. From each Bayesian posterior,
the parameter estimates regarding mutation rates (including clock rate)
were obtained and compared to determine which realized mutation rate is lower.
The Markov chain Monte Carlo was set up in such a way that the effective sample
size for the likelihood of the inference model is above the recommended value
of 200 (\cite{bouckaert2014beast}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{MHC-I}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{MHC-II}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evolutionary conservation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
